---
title: "Drought Deficit"
format: html
editor_options: 
  chunk_output_type: console
---

```{r setup}
#| echo: false
#| hide: true
#| warning: false
#| message: false
#| results: false

library(here)
source(here("code/f_get_fonts.R"))

# set fonts depending on system:
font1 <- if(get_os()=="windows"){
  "Century Gothic"
} else {
  "Questrial"
}
font2 <- "Roboto Condensed"

# load fonts
f_get_fonts(font1, font2)

# now load libraries
library(tidyverse)
library(here)
#library(reticulate)
library(sf)
library(tigris)
options(tigris_use_cache = TRUE)
library(ggiraph)
library(glue)
library(fs)
library(patchwork)
library(mapview)
mapviewOptions(fgb=FALSE)


```


## Drought Subsurface Deficit Analysis

The analysis is based on a number of papers, but see [here.](https://hess.copernicus.org/articles/25/2861/2021/)^[Dralle, D. N., W. J. Hahm, D. M. Rempe, N. J. Karst, S. E. Thompson, and W. E. Dietrich. 2018. Quantification of the seasonal hillslope water storage that does not drive streamflow. Hydrological processes 32:1978–1992.]

### Code 

The python code [found here](https://colab.research.google.com/drive/1Fijggch1V3wcV8iU2NnlVVm77PGMd_ov?usp=sharing#scrollTo=tM-RO4DldBbR) runs the analysis required to generate the subsurface water deficit.

```{bash}
#| eval: false
#| echo: false
 
!pip install -q geopandas 

%matplotlib inline
import os
import geopandas as gp
import sys
import seaborn as sns
from matplotlib import pyplot as plt
import numpy as np
from google.colab import files
import pandas as pd
import scipy
import glob
import datetime
from shapely.geometry import Point
import requests
import ee
from google.colab import files
from shapely.geometry import Polygon, MultiPolygon
import ee
%config InlineBackend.figure_format = 'retina'

ee.Authenticate()

# specify point locations and site names for deficit calculations
pts = [Point( -119.188446,37.063012),
       Point(-119.258238,37.029470),
       Point( -123.630644,39.723181),
       Point(-123.479911,39.566365)]

# specify site names
sites = ['PROV','SOAP','ELDER','DRY']

# geodataframe to save data
gdf = gp.GeoDataFrame({'site':sites},geometry=pts,crs='epsg:4326',index=[0,1,2,3])

# Create earth engine feature collection
fts_list = []
geos = []
names = []
for i,row in gdf.iterrows():
  x = row.geometry.centroid.x
  y = row.geometry.centroid.y
  geo = ee.Geometry.Point((x,y))
  geos.append(geo)
  names.append(row.site)
  ft = ee.Feature(geo)
  ft = ft.set('site',row.site)
  fts_list.append(ft)
fts = ee.FeatureCollection(fts_list)

# Start/stop time of analysis...PML v2 only goes through 2021
start='2000-10-01'
stop='2021-10-01'

#Flux is stored as 100*mm/day, and duration between samples is 8 days
multiplier = 8.0 
dataset = ee.ImageCollection("projects/pml_evapotranspiration/PML/OUTPUT/PML_V2_8day_v016");
scale = 0.01

#### calculating F_out WITHOUT SNOW CORRECTION ####
# mapper for getting single ET raster for each timestep
def sumpml(image):
  # ET bands into one band
  temp = image.select('Es').add(image.select('Ec')).add(image.select('Ei')).multiply(scale)

  #get the first band of this new temporary image, rename it to 'ET', 
  #reproject it, and then multiply by multiplier
  temp = temp.select([0], ['ET']).reproject(
      crs='EPSG:4326',scale=1000
      ).multiply(multiplier)

  #temp should now be total ET in mm over the time window between images
  #assign the datetime stamp and the index from the original image
  temp = temp.set('system:time_start', image.get('system:time_start'))
  temp = temp.set('system:index', image.get('system:index'))
  return temp

#now, actually map the pml image collection w/ this function,
# in order to make a new combined ET image collection
et = ee.ImageCollection(dataset.map(sumpml).select('ET')).filter(ee.Filter.date(start, stop))
et = et.toBands()

# Use PRISM rainfall and PML ET to calculate deficit at monthly timestep
data = {}
for i,geo in enumerate(geos):
  ft = et.reduceRegion(ee.Reducer.mean(), geo, 1000)
  vals = ft.getInfo()
  dates = pd.to_datetime([item.split('_')[0] for item in list(vals)])
  etvals = [item for item in vals.values()]
  etdf = pd.DataFrame({'ET':etvals},index=dates)
  etdf = etdf.resample('M').sum()

  dataset = ee.ImageCollection("OREGONSTATE/PRISM/AN81m").filter(ee.Filter.date(start, stop))
  ppt = dataset.select('ppt').toBands()
  ft = ppt.reduceRegion(ee.Reducer.mean(), geo, 4000)
  vals = ft.getInfo()
  dates = pd.to_datetime([item.split('_')[0] for item in list(vals)],format='%Y%m')
  pptvals = [item for item in vals.values()]
  pptdf = pd.DataFrame({'P':pptvals},index=dates).resample('M').sum()
  
  # Combine
  df = pd.concat([etdf, pptdf],axis=1)
  df['wy'] = [item.year if item.month<10 else item.year+1 for item in df.index]
  df['sumP'] = df.groupby('wy')['P'].cumsum()

  deficit = [0]
  counter = 0
  # Wang-Erlandsson deficit calculation
  for idx,row in df.iterrows():
    A = row.ET - row.P
    next = np.max([0, deficit[counter] + A])
    deficit.append(next)
    counter = counter + 1
  df['D'] = deficit[1:]
  data[names[i]] = df
  
for site in list(data):
  tempdf = data[site]
  tempdf.to_csv('./%s_PML_PPT.csv'%site)
  
# Download data files for each site
# can also just grab from file directory folder icon on left side
# you will have to "allow" colab to download multiple files
# via your browser, most likely
for site in list(data):
  files.download('./%s_PML_PPT.csv'%site) 
  
f,axs = plt.subplots(1,len(data),figsize=(3*len(data),2.5),sharey=True)
axs[0].set_ylabel('Deficit (mm)')
for i,site in enumerate(list(data)):
  ax = axs[i]
  temp = data[site]
  temp.D.plot(ax=ax,label=site)
  ax.set_title(site)
  ax.set_xlim('2005-10','2021-10')
f.autofmt_xdate()
f.tight_layout()
```

### Sites

Here's a sampling of sites across CA that represent a number of different Coastal and Sierran Rivers, and incidentally, different subsurface storage sizes.

### Plotting Deficit

Here we quickly plot these sites with mapview to get a sense of where things are.

```{r}
#| echo: false
#| eval: true


# make a tibble
sites <- tibble(sites=c("PROV", "SOAP", "ELDER","ALAMEDA", "SCARP","NFA", "DRY"),
           lon = c(-119.188446,-119.258238, -123.630644,-121.79203,-121.26971,-120.92493, -123.479911),
           lat = c(37.063012, 37.029470,39.723181, 37.50237, 35.82157, 39.10509, 39.566365))

# make sf
sites_sf <- sites %>% 
  st_as_sf(coords=c("lon","lat"), remove=FALSE, crs=4326)

mapview(sites_sf, zcol="sites", layer.name="Sites")

```

## Deficit Data

These data were calculated for 2001-2021. Data are pulled in and cleaned, compiled, and then plotted below.

```{r}
#| echo: false
#| eval: true
#| message: false
#| results: hide

# get list of files
library(fs)
pml_files <- fs::dir_ls(here("data_out"), regexp = "*.csv")
# get data
df <- read_csv(pml_files, id = "sitepath") %>%
  rename("date"=2) %>% 
  mutate(site = path_ext_remove(path_file(sitepath)),
         site = gsub("_PML_PPT","", x = site))

# get county data
# now a map
cnty <- counties(state = "CA")

```

And plot!

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false

# all together
gg1 <- ggplot() + 
  geom_line_interactive(data=df, 
                        aes(x=date, y=D, color=site, tooltip = site,
      data_id = site)) +
  labs(x="", y="Deficit (mm)")+
  scale_color_viridis_d("")+
  theme_classic(base_family = fnt_header)+
    theme(legend.position = c(0.1, 0.82),
          legend.text = element_text(size=7, family = fnt_text))


# map
ggmap1 <- ggplot() +
  geom_sf(data=cnty, color="gray", linewidth=0.5)+
  geom_sf_interactive(
    data=sites_sf,
    aes(
      fill=sites,
      tooltip = sites,
      data_id = sites
    ), show.legend=FALSE, size=3, pch=21,stroke=0.7, alpha=0.7, color="white") +
  scale_fill_viridis_d("") +
  theme_void()
ggmap1

# plot static
plot_def_stat <- gg1 + 
  inset_element(ggmap1, left = 0.7, bottom = 0.7, right = 1, top = 1) + 
  #plot_layout(widths = c(2, 1)) + 
  plot_annotation(title="Subsurface Deficit", 
                  caption="based on Dralle et al., data 2000-2021", 
                  theme = theme(plot.title = element_text(size = 16)))

plot_def_stat
ggsave(plot_def_stat, filename = "figs/subsurface_water_deficit_sample_sites.png", width = 11, height = 8, dpi=300, bg="white")

# plot dynamic
plot_def_int <- girafe(ggobj = gg1 + inset_element(ggmap1, left = 0.7, bottom = 0.7, right = 1, top = 1) + 
              #plot_layout(widths = c(2, 1)) + 
              plot_annotation(title="Subsurface Deficit", 
                              caption="based on Dralle et al., data 2000-2021", 
                              theme = theme(plot.title = element_text(size = 16))), 
            width_svg = 8, height_svg = 6,
            options = list(
              opts_zoom = opts_zoom(min = 1, max = 4),
              opts_hover_inv(css = "opacity:0.1;"),
              opts_hover(css = "stroke-width:2;")
            ))

plot_def_int

```

And a faceted version here.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false

gg2 <- ggplot() + 
  geom_line(data=df, aes(x=date, y=D, color=site), show.legend=FALSE) +
  labs(x="", y="Deficit (mm)")+
  scale_color_viridis_d("Site")+
  theme_classic(base_family = fnt_header) + 
  facet_grid(site~.)

(plot_def_stat2 <- gg2 + ggmap1 + 
  plot_layout(widths = c(1, 0.25)) + 
  plot_annotation(title="Subsurface Deficit", 
                  caption="based on Dralle et al., data 2000-2021", 
                  theme = theme(plot.title = element_text(size = 16))))

ggsave(plot_def_stat2, filename = "figs/subsurface_water_deficit_sample_sites_facet.png", width = 11, height = 8, dpi=300, bg="white")


```


## Zoom to a specific Time Period

Let's look at the 2014+ drought.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false

df_zoom <- df %>% filter(date > ymd("2013-10-01"),
                         !is.na(D),
                         !site %in% c("PROV","DRY"))

sites_sf_trim <- sites_sf %>% filter(sites %in% c("ALAMEDA","ELDER","NFA","SCARP","SOAP"))

# map
(ggmap2 <- ggplot() +
  geom_sf(data=cnty, color="gray", linewidth=0.5)+
  geom_sf_text(data=sites_sf_trim, aes(label=sites), nudge_x = 0.2, nudge_y = 0.5, size=1.4)+
  geom_sf_interactive(
    data=sites_sf_trim,
    aes(
      fill=sites,
      tooltip = sites,
      data_id = sites
    ), show.legend=FALSE, size=5, pch=21,stroke=0.7, alpha=0.7, color="white") +
  scale_fill_viridis_d("") +
  theme_void())

gg3 <- ggplot() + 
  geom_line(data=df_zoom, aes(x=date, y=D, color=site), show.legend=TRUE) +
  labs(x="", y="Deficit (mm)")+
  scale_x_date(date_breaks = "6 months",date_labels = "%b-%y")+
  scale_color_viridis_d("Site")+
  cowplot::theme_half_open(font_family = fnt_header) + 
  cowplot::background_grid("x")+
  theme(axis.text.x = element_text(vjust=0.5, hjust=0.5, size=8, angle=90))

(plot_def_stat3 <- gg3 + inset_element(ggmap2, left = 0.65, bottom = 0.65, right = 1, top = 1) +  
  plot_annotation(title="Subsurface Deficit", 
                  caption="based on Dralle et al., data 2013-2021", 
                  theme = theme(plot.title = element_text(size = 16))))

ggsave(plot_def_stat3, filename = "figs/subsurface_water_deficit_sample_sites_facet_drought.png", width = 11, height = 8, dpi=300, bg="white")


```

