---
title: "Drought Deficit"
format: html
editor_options: 
  chunk_output_type: console
---

```{r setup}
#| echo: false
#| hide: true
#| warning: false
#| message: false
#| results: false

library(here)
source(here("code/f_get_fonts.R"))

# set fonts depending on system:
font1 <- if(get_os()=="windows"){
  "Century Gothic"
} else {
  "Atkinson Hyperlegible"
}
font2 <- "Roboto Condensed"

# load fonts
f_get_fonts(font1, font2)

# now load libraries
library(tidyverse)
library(here)
#library(reticulate)
library(sf)
library(mapview)
mapviewOptions(fgb=FALSE)


```


## Drought Subsurface Deficit Analysis

Based on Dralle et al (numerous) papers.

### Code 

Code is based on this analysis in `python`.

```{bash}
#| eval: false
#| echo: true
#| code-line-numbers: "|24"
!pip install -q geopandas 

%matplotlib inline
import os
import geopandas as gp
import sys
import seaborn as sns
from matplotlib import pyplot as plt
import numpy as np
from google.colab import files
import pandas as pd
import scipy
import glob
import datetime
from shapely.geometry import Point
import requests
import ee
from google.colab import files
from shapely.geometry import Polygon, MultiPolygon
import ee
%config InlineBackend.figure_format = 'retina'

ee.Authenticate()
```

### Sites

As a test, we used the following locations:

 - PROV = -119.188446,37.063012
 - ELDER = -123.630644,39.723181
 - ALAMEDA = -121.79203, 37.50237
 - SCARP = -121.26971, 35.82157
 - NFA = -120.92493, 39.10509

These represent a number of different Coastal and Sierran Rivers. 

### Plotting Deficit

Here we quickly plot these sites:

```{r}

# make a tibble
sites <- tibble(sites=c("PROV", "ELDER","ALAMEDA", "SCARP","NFA"),
           lon = c(-119.188446, -123.630644,-121.79203,-121.26971,-120.92493),
           lat = c(37.063012, 39.723181, 37.50237, 35.82157, 39.10509))

# make sf
sites_sf <- sites %>% 
  st_as_sf(coords=c("lon","lat"), remove=FALSE, crs=4326)

mapview(sites_sf, zcol="sites")

```

## Deficit Data

These data were calculated for 2001-2021.

```{r}
#| echo: false
#| eval: true
#| message: false


# get list of files
library(fs)
pml_files <- fs::dir_ls(here("data_out"), regexp = "*.csv")
# get data
df <- read_csv(pml_files, id = "sitepath") %>%
  rename("date"=2) %>% 
  mutate(site = path_ext_remove(path_file(sitepath)),
         site = gsub("_PML_PPT","", x = site))



```

And plot!

```{r}
#| echo: false
#| eval: true
#| message: false


ggplot() + 
  geom_line(data=df, aes(x=date, y=D, color=site)) +
  labs(x="", y="Deficit (mm)")+
  scale_color_viridis_d("Site")+
  theme_classic(base_family = "Atkinson Hyperlegible")
```


```{r}
#| echo: false
#| eval: true
#| message: false


ggplot() + 
  geom_line(data=df, aes(x=date, y=D, color=site), show.legend=FALSE) +
  labs(x="", y="Deficit (mm)")+
  scale_color_viridis_d("Site")+
  theme_classic(base_family = "Atkinson Hyperlegible") + 
  facet_grid(site~.)
```
